Paradigms in Physics 2.0

### 2017-06-21 paper.tex
add history and new courses text (the latter from the grant proposal)

%\documentclass[english,aps,prstper,reprint,showpacs]{revtex4-1}   % This version should work (prstper option), but actually formats reference numbers incorrectly.
\documentclass[english,aps,pra,reprint,noshowpacs,superscriptaddress]{revtex4-1}   
% Using pra instead of prstper gives correct square-bracket (not superscript) reference formatting
\usepackage[T1]{fontenc}	% should generally be included for better accented-word behavior
\usepackage[latin9]{inputenc}	% should generally be included for better accent behavior
\usepackage{geometry}		% for controlling page margins
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=0.75in,rmargin=0.75in}	% define margins
\usepackage{graphicx}
\usepackage[above,below]{placeins}	% allows use of \FloatBar­rier command to force section barriers
\usepackage{times}
% Next three lines are optional, use the hyperref package to make URLs and reference links live.
\usepackage{hyperref}  
\hypersetup{colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue}   
\urlstyle{same}
\pagestyle{empty}			% page numbers added later, when compiling the whole proceedings
\begin{document}

@@

\section{Introduction}
The Paradigms in Physics program

\subsection{Background}
\emph{What is the Paradigms now?}
\subsection{History of the Paradigms}
\emph{How did we get here?~\cite{manogue2001paradigms}}
\subsection{Motivation for the change}
There were a number of factors that motivated us to embark on the
Paradigms~2.0 process.  The original Paradigms in Physics curricular
reform from 200? was showing its age.  In the intervening years we
have made a large number of modifications, such as reordering
Paradigms, introducing one new Paradigm, and introducing and changing
computational physics courses.

@@

\subsection{Sophomore courses}
We introduced two new sophomore-level courses: \emph{Challenges in
  Contemporary Physics}, and \emph{Theoretical Mechanics}.  Both of
these courses are designed to ramp up student mathematical abilities
prior to their junior year.  The Challenges course will focus on
estimation, dimensional reasoning, and interpretation of integrals,
while Theoretical Mechanics will teach power series approximations,
and expose students to increased levels of mathematical
sophistication.

@@

\subsubsection{Challenges in Contemporary Physics}
\emph{Course overview: relevant to world, benefit for diversity and
  recruitment, replaces Modern Physics. Topics.}
\subsubsection{Techniques of Theoretical Mechanics}
\emph{Course overview: special relativity, rocket motion, Lagrangian
  mechanics.}

@@

\begin{table*}[htbp]
\caption{Hypothetical student schedule for a student entering without
  calculus.  Students with AP Calculus can take the junior-year
  courses one year earlier, and students who start calculus-based
  physics in the Fall of their second year are not
  delayed.\label{schedule}}
\begin{ruledtabular}
\begin{tabular}{rlll}
  \textbf{Year} & \textbf{Fall} & \textbf{Winter} & \textbf{Spring} \\
  \hline
  1 & \mathcourse{Differential calculus}{251} &
  \mathcourse{Integral calculus}{252} &
  \mathcourse{Multivariable calculus}{254} \\
    & & \onecredit{Freshman Seminar} & \fourcredit{Calc Physics I} \\
\hline  2 & \mathcourse{Vector calculus}{255}
  & \mathcourse{Differential equations}{256}
  & \mathcourse{Linear algebra}{341} \\
  & \fourcredit{Calc Physics II} & \fourcredit{Calc Physics III} &\mathcourse{Series \&
    sequences}{253} \\
  && \noted{Challenges}{3} & \noted{Theoretical Mechanics}{3}
  \\
\hline 3 & \paradigm{Quantum fundamentals} & \paradigm{Osciallations \& waves} & \paradigm{Static fields}
\\
  & \paradigm{Energy \& entropy} & \paradigm{Periodic systems} & \paradigm{Central forces}
\\
  & \onecredit{Computational lab I} & \onecredit{Computational lab II} & \onecredit{Computational lab III}
\\
  & \threecredit{Electronics} & \threecredit{Computer interfacing} & \onecredit{Research}
\\
\hline 4 & \capstone{Electromagnetism capstone} & \capstone{Quantum capstone} & \capstone{Thermal capstone}
\\
& \onecredit{Thesis} & \onecredit{Thesis} & \onecredit{Thesis}
\\
& \onecredit{Research} & \onecredit{Research}
\end{tabular}
\end{ruledtabular}
\end{table*}

\begin{table*}[htbp]
\caption{Hypothetical transfer student schedule.  Most transfer
  students will have a heavier schedule in their transfer year due
  to need to catch up on required math courses.\label{schedule}}
\begin{ruledtabular}
\begin{tabular}{rlll}
  \textbf{Year} & \textbf{Fall} & \textbf{Winter} & \textbf{Spring} \\
  \hline
3 & \paradigm{Quantum fundamentals} & \paradigm{Osciallations \& waves} & \paradigm{Static fields}
\\
  & \paradigm{Energy \& entropy} & \paradigm{Periodic systems} & \paradigm{Central forces}
\\
  & \onecredit{Computational lab I} & \onecredit{Computational lab II} & \onecredit{Computational lab III}
\\
  & \threecredit{Electronics} & \threecredit{Computer interfacing} & \onecredit{Research}
\\
& \mathcourse{Linear algebra}{341} & \noted{Challenges}{3} & \noted{Theoretical Mechanics}{3}
\\
&&\mathcourse{Vector calculus}{255}
\\
\hline 4 & \capstone{Electromagnetism capstone} & \capstone{Quantum capstone} & \capstone{Thermal capstone}
\\
& \onecredit{Thesis} & \onecredit{Thesis} & \onecredit{Thesis}
\\
& \onecredit{Research} & \onecredit{Research}
\end{tabular}
\end{ruledtabular}
\end{table*}

>>>>

%\documentclass[english,aps,prstper,reprint,showpacs]{revtex4-1}   % This version should work (prstper option), but actually formats reference numbers incorrectly.
\documentclass[english,aps,pra,reprint,noshowpacs,superscriptaddress]{revtex4-1}   
% Using pra instead of prstper gives correct square-bracket (not superscript) reference formatting
\usepackage[T1]{fontenc}	% should generally be included for better accented-word behavior
\usepackage[latin9]{inputenc}	% should generally be included for better accent behavior
\usepackage{geometry}		% for controlling page margins
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=0.75in,rmargin=0.75in}	% define margins
\usepackage{graphicx}
\usepackage[above,below]{placeins}	% allows use of \FloatBarÂ­rier command to force section barriers
\usepackage{times}
% Next three lines are optional, use the hyperref package to make URLs and reference links live.
\usepackage{hyperref}  
\hypersetup{colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue}   
\urlstyle{same}
\pagestyle{empty}			% page numbers added later, when compiling the whole proceedings
\begin{document}

@@

\section{Introduction}
The Paradigms in Physics program began around the time when our
current students were born~\cite{manogue2001paradigms}.  The time has
arrived for a change.  Over the years, we have identified a number of
problems in our curriculum that we were unable to address without
addressing the physics major curriculum as a whole.  In this paper, we
share the motivation for our Paradigms 2.0 effort, the process by
which we achieved consensus on an improved curriculum, and will give
a summary and a few highlights of the new program.

\subsection{Background}
The Paradigms in Physics project began in 1996, when three faculty
members at Oregon State applied to the NSF for funding to redesign the
upper-division curriculum.  The motivation for the original Paradigms
project was in fact similar to the motivation for Paradigms 2.0: a
desire to soften the ``brick wall'' encountered by students upon
reaching the junior year, within the constraint that transfer students
must be able to graduate with two years of upper-division
courses~\cite{manogue2001paradigms}.  The process of redesigning the
curriculum was dominated by the creation index cards listing subject
content, and sorting these cards into courses.  This process involved
the entire faculty, and culminated in a unanimous agreement to adopt
the resulting curriculum.

The resulting curriculum was primarily composed of junior-year
``Paradigms'' courses, followed by senior-year ``Capstone'' courses.
The Paradigms were intensive 2-credit, 3-week courses which meet every
day, for a total of seven hours per week.  Paradigms incorporate
laboratory experiences and active engagement into the class, and
typically have two problem sets per week.  Capstones were more
traditional 3-credit courses which meet three hours per week for an
entire quarter.  The Paradigms courses had integrated math content,
followed by a Math Methods course.

A coupe of decades have passed since the original Paradigms effort.
During this time we have made a number of changes: e.g. a new
\emph{first} Paradigm was introduced to soften the beginning of the
junior year, the order of Paradigms was shuffled more than once, Math
Methods was moved from the beginning of the senior year to the end of
the junior year, and a computational laboratory course was introduced
to accompany the Paradigms.  We maintained the tradition of meeting
every three weeks to discuss issues relating to upper-division
teaching.  New faculty arrived and learned to teach the new courses,
and introduced their own approach to the courses.

\subsection{Motivation for Paradigms 2.0}
There were a number of factors that motivated us to embark on the
Paradigms~2.0 process.  The original Paradigms in Physics curricular
reform from 200? was showing its age.  In the intervening years we
have made a large number of modifications, such as reordering
Paradigms, introducing one new Paradigm, and introducing and changing
computational physics courses.

@@

\subsection{Sophomore courses}
We introduced two new sophomore-level courses: \emph{Challenges in
  Contemporary Physics}, and \emph{Theoretical Mechanics}.  Both of
these courses ramp up student mathematical abilities prior to their
junior year.  The Challenges course will focus on estimation,
dimensional reasoning, and interpretation of integrals, while
Theoretical Mechanics will teach power series approximations, and
expose students to increased levels of mathematical sophistication.

@@

\subsubsection{Challenges in Contemporary Physics}
Climate change and renewable energy are major contemporary
challenges. Students are inspired to engage with these challenges, and
physics will play a critical role in addressing these challenges.
Former Energy Secretary Steve Chu recently noted that ``physics majors
seem to be increasing,'' and pointed to millennials' interests in the
risks of climate change, clean energy, and
sustainability~\cite{kramer2016gathering}.  However, most majors get
little or no interaction with such challenges in their physics
classes.

We have introduced a course that fits seamlessly into the physics
major sequence, that focuses on the contemporary challenges of
sustainable energy and climate change. The course will be a bridge
between introductory physics and upper-division classes for both
physics majors and students from related disciplines such as chemistry
and engineering. The challenges theme will motivate the introduction
of thermal physics and quantum mechanics as well as the development of
physical reasoning skills. We anticipate that an emphasis on the role
of physics in overcoming societal challenges will help us recruit and
retain a greater number of physics majors, especially women and
under-represented minorities.

At Oregon State, as at many other universities, a course in Modern
Physics traditionally bridges the lower and upper divisions and serves
as an introduction to quantum mechanics. Such courses typically follow
a historical narrative, celebrating the achievements of men like
Planck, Einstein, Rutherford, Bohr, Bragg, Millikan, de Broglie,
Pauli, Schroedinger, and Dirac. Indeed, it is a wonderful and exciting
story - and inspiring for those who imagine themselves as
reincarnations of these men. However, we suggest that this historical
narrative does not attract a wide cross-section of participants to
physics. Instead, we have introduced a forward-looking narrative that
emphasizes the relevance of physics today unconstrained by the details
of the great transition from 1905 to 1945.

Each physics topic is linked to contemporary challenges. An
understanding of electromagnetic waves is required to understand how
light interacts with vibrational modes of CO$_2$. The equipartition
theorem explains the heat energy stored in the ocean. Quantum
mechanics explains the blackbody spectrum, semiconductor electronic
structure, and solar photovoltaics. The Carnot cycle convinces us that
a 95\%-efficient gas furnace cannot compete with an efficient heat
pump.

The new perspectives associated with this course, along with an
emphasis on expanding students' reasoning skills, will strengthen the
connection between lower- and upper-division physics classes. Students
will learn the importance of approximate calculations (Fermi
estimates) that establish reasonable bounds on the exact
answer. Students will learn that equations help us interpret the
physical world, i.e. equations are used for much more than mindless
algebraic manipulations.  Students will learn how integration and
differentiation can be applied to real data - not just the perfect
functions that are taught in math class. Students will learn how
graphing of relationships is key tool for physicists. The utility of
these skills will be firmly established by application to contemporary
challenges.

\subsubsection{Techniques of Theoretical Mechanics}

To physicists, mathematics is a language for telling stories about
conceptual physical relationships.  Our understandings of conceptual
connections guide us when we manipulate
equations~\cite{gire2008resources, bing2007cognitive,
  wittmann2015mathematical}[75].  For example, when solving ordinary
differential equations by separating variables, we choose limits of
integration so that the lower (and upper) limits for all variables
describe the same physical state.  We monitor algebraic manipulations
using conceptual understandings of physics quantities. For example, we
make sure that each term in a sum has the same physical dimensions
(e.g. you can't add a frequency to a frequency squared). We make sure
that expressions on both sides of an equals sign have the same
dimensions~\cite{lenz2016dimensional}. We consider limiting cases to
make sure that a derived expression reasonably describes the system in
extreme states~\cite{singh2002physical}. Our geometric understandings
guide the formulation of equations~\cite{gire2014arrows}. Sometimes,
we do not have adequate mathematics to describe complex physical
systems. We use conceptual understandings to make choices about
reasonable idealizations in order to create useful mathematical
models~\cite{bing2007cognitive}.

Being fluent in this physics language involves both (1) translating a
conceptual story into a mathematical representation and (2) discerning
a conceptual story from a mathematical
representation~\cite{tuminaro2007elements}. At the introductory level,
students often see concepts and mathematics as being separate. These
students do not expect equations to help their conceptual
understanding~\cite{adams2006new, redish1998expectations} and have
trouble formulating equations from a conceptual analysis. Students who
do not have mathematical fluency for physics struggle in advanced
physics courses~\cite{manogue2009cognitive}.

We propose a course that explicitly focuses on physics majors'
mathematical fluency and sense- making skills. Introductory courses
begin to develop these aspects, but the primary goals of those courses
tend to be for students to apply simple concepts (Newton's Laws,
energy conservation, etc.)  to relatively straightforward and very
idealized problems.

This course will be taken just after the introductory physics
sequence. It will leverage students' understandings of simple
mechanical systems from introductory physics to develop their
mathematical fluency in the context of less idealized, more
sophisticated systems. Each physics topic (velocity-dependent forces,
Lagrangian mechanics, Hamiltonian mechanics, special relativity) will
focus on specific math methods (e.g. differential equations, calculus
of variations, generalized coordinates, Legendre transformations,
linear transformations, hyperbola geometry) and will explicitly
feature strategies to make physical sense of mathematics (symbolic
problem solving, reading equations like sentences, making
idealization, coordinating representations, limiting case analysis,
dimensional checking).

@@

%% \begin{table*}[htbp]
%% \caption{Hypothetical student schedule for a student entering without
%%   calculus.  Students with AP Calculus can take the junior-year
%%   courses one year earlier, and students who start calculus-based
%%   physics in the Fall of their second year are not
%%   delayed.\label{schedule}}
%% \begin{ruledtabular}
%% \begin{tabular}{rlll}
%%   \textbf{Year} & \textbf{Fall} & \textbf{Winter} & \textbf{Spring} \\
%%   \hline
%%   1 & \mathcourse{Differential calculus}{251} &
%%   \mathcourse{Integral calculus}{252} &
%%   \mathcourse{Multivariable calculus}{254} \\
%%     & & \onecredit{Freshman Seminar} & \fourcredit{Calc Physics I} \\
%% \hline  2 & \mathcourse{Vector calculus}{255}
%%   & \mathcourse{Differential equations}{256}
%%   & \mathcourse{Linear algebra}{341} \\
%%   & \fourcredit{Calc Physics II} & \fourcredit{Calc Physics III} &\mathcourse{Series \&
%%     sequences}{253} \\
%%   && \noted{Challenges}{3} & \noted{Theoretical Mechanics}{3}
%%   \\
%% \hline 3 & \paradigm{Quantum fundamentals} & \paradigm{Osciallations \& waves} & \paradigm{Static fields}
%% \\
%%   & \paradigm{Energy \& entropy} & \paradigm{Periodic systems} & \paradigm{Central forces}
%% \\
%%   & \onecredit{Computational lab I} & \onecredit{Computational lab II} & \onecredit{Computational lab III}
%% \\
%%   & \threecredit{Electronics} & \threecredit{Computer interfacing} & \onecredit{Research}
%% \\
%% \hline 4 & \capstone{Electromagnetism capstone} & \capstone{Quantum capstone} & \capstone{Thermal capstone}
%% \\
%% & \onecredit{Thesis} & \onecredit{Thesis} & \onecredit{Thesis}
%% \\
%% & \onecredit{Research} & \onecredit{Research}
%% \end{tabular}
%% \end{ruledtabular}
%% \end{table*}

%% \begin{table*}[htbp]
%% \caption{Hypothetical transfer student schedule.  Most transfer
%%   students will have a heavier schedule in their transfer year due
%%   to need to catch up on required math courses.\label{schedule}}
%% \begin{ruledtabular}
%% \begin{tabular}{rlll}
%%   \textbf{Year} & \textbf{Fall} & \textbf{Winter} & \textbf{Spring} \\
%%   \hline
%% 3 & \paradigm{Quantum fundamentals} & \paradigm{Osciallations \& waves} & \paradigm{Static fields}
%% \\
%%   & \paradigm{Energy \& entropy} & \paradigm{Periodic systems} & \paradigm{Central forces}
%% \\
%%   & \onecredit{Computational lab I} & \onecredit{Computational lab II} & \onecredit{Computational lab III}
%% \\
%%   & \threecredit{Electronics} & \threecredit{Computer interfacing} & \onecredit{Research}
%% \\
%% & \mathcourse{Linear algebra}{341} & \noted{Challenges}{3} & \noted{Theoretical Mechanics}{3}
%% \\
%% &&\mathcourse{Vector calculus}{255}
%% \\
%% \hline 4 & \capstone{Electromagnetism capstone} & \capstone{Quantum capstone} & \capstone{Thermal capstone}
%% \\
%% & \onecredit{Thesis} & \onecredit{Thesis} & \onecredit{Thesis}
%% \\
%% & \onecredit{Research} & \onecredit{Research}
%% \end{tabular}
%% \end{ruledtabular}
%% \end{table*}
